---
title: "NYPD Shooting incident analysis"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Source ###

The data is from https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic.
It lists of every shooting incident that occurred in New York City going back to 2006
through the end of the previous calendar year (2022). According to the website, This
data is manually extracted every quarter and reviewed by the Office of Management
Analysis and Planning before being posted on the NYPD website.

### Import the NYPD shooting incident data

```{r import_csv, echo=TRUE}
library(tidyverse)
library(lubridate)
filename = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd <- read_csv(filename)
nypd
```

### Transform the data

1) Merge OCCUR_DATE and OCCUR_TIME into "dttm" format:
```{r merge_date_time, echo=TRUE}
nypd <- nypd %>%
    unite(datetime, 2:3, sep = " ", remove = FALSE) %>%
    mutate(
        datetime = as.POSIXct(datetime, format = "%m/%d/%Y %H:%M:%S")) %>%
    ungroup()
nypd %>%
    select(datetime, OCCUR_DATE, OCCUR_TIME, everything())
```

2) Keep only the columns we are interested in for this analysis:
```{r keep_needed_columns, echo=TRUE}
nypd <- nypd %>%
    select(datetime, BORO, LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)
```


### Analyze the data

#### 1) Check the total number of incidents in each borough
```{r get_boro_stat, echo=TRUE}
boro_stat <- nypd %>%
    group_by(BORO) %>%
    summarise(count=n()) %>%
    select(BORO, count) %>%
    ungroup()
boro_stat
```

```{r draw_boro_stat_pie, echo=TRUE}
slices <- boro_stat$count
boro <- boro_stat$BORO
percent <- round(slices/sum(slices)*100)
boro <- paste(boro, percent)
boro <- paste(boro, "%", sep="")
pie(slices, labels=boro, main="Pie Chart of Boroughs")
```

As we can see, Brooklyn had the highest number of shooting incidents, accounting
for 40% of all cases, followed by the Bronx, which recorded 29% of the incidents.

#### 2) Check the number of incidents over the years


```{r generate_year_stat, echo=TRUE}
nypd_year <- nypd %>%
    mutate(year = format(datetime, "%Y")) %>%
    group_by(year) %>%
    summarise(count=n()) %>%
    select(year, count) %>%
    ungroup()
nypd_year
```
```{r draw_year_stat, echo=TRUE}
nypd_year %>%
  ggplot(aes(x=year, y=count, fill=count)) +
  geom_bar(width=0.8, stat="identity") +
  labs(title = "NY shooting incidents over the years", y = NULL)
```

From 2006 to 2019, the situation showed improvement, particularly from 2017 to 2019
when the annual number of incidents remained below 1000. However, in 2020, the figure
almost doubled compared to the previous year, likely influenced significantly by
the COVID-19 pandemic.


#### 3) Check the number of incidents for each hour of the day

```{r generate_hour_stat, echo=TRUE}
nypd_hour_stat <- nypd %>%
    mutate(hour = format(datetime, "%H")) %>%
    group_by(hour) %>%
    summarise(count=n()) %>%
    select(hour, count) %>%
    ungroup()
print(nypd_hour_stat, n = Inf)
```

```{r draw_hour_stat, echo=TRUE}
nypd_hour_stat %>%
  ggplot(aes(x=hour, y=count, fill=count)) +
  geom_bar(width=0.8, stat="identity") +
  labs(title = "NY shooting incidents over each hour of the day", y = NULL)
```

As showed above, the period with the highest number of shooting incidents was
between 23:00 and 00:00.


#### 4) Check incidents for different types of locations

By checking the data first, seems lots of cases did not have exact location information.

```{r check_loc_class_data, echo=TRUE}
nypd %>%
    group_by(LOC_CLASSFCTN_DESC) %>%
    summarise(count=n()) %>%
    select(LOC_CLASSFCTN_DESC, count) %>%
    ungroup()
```

So I have to drop these NAs first before generating statistical data:

```{r get_loc_class_data, echo=TRUE}
nypd_loc_class <- nypd %>%
    drop_na(LOC_CLASSFCTN_DESC) %>%
    group_by(LOC_CLASSFCTN_DESC) %>%
    summarise(count=n()) %>%
    select(LOC_CLASSFCTN_DESC, count) %>%
    ungroup()
nypd_loc_class
```

```{r draw_loc_class_pie, echo=TRUE}
slices <- nypd_loc_class$count
loc_class <- nypd_loc_class$LOC_CLASSFCTN_DESC
percent <- round(slices/sum(slices)*100)
loc_class <- paste(loc_class, percent)
loc_class <- paste(loc_class, "%", sep="")
pie(slices, labels=loc_class, main="Pie Chart of Location class", cex=0.8)
```

The pie chart shows more than half the incidents happened in the street, followed by
the second-highest in residential houses!


#### 5) Retrieve statistical data based on the gender of the victims

```{r get_victim_gender, echo=TRUE}
nypd_victim_gender <- nypd %>%
    group_by(VIC_SEX) %>%
    summarise(count=n()) %>%
    select(VIC_SEX, count) %>%
    ungroup()
nypd_victim_gender
```

```{r draw_victim_gendor_pie, echo=TRUE}
slices <- nypd_victim_gender$count
vic_sex <- nypd_victim_gender$VIC_SEX
percent <- round(slices/sum(slices)*100)
vic_sex <- paste(vic_sex, percent)
vic_sex <- paste(vic_sex, "%", sep="")
pie(slices, labels=vic_sex, main="Pie Chart of the gender of the victim")
```

As showed above, the majority of shooting victims were male, constituting 90% of
the cases.

### Create model for better understanding the data

1) Generate the male victim incidents for each year

```{r get_male_vic_data, echo=TRUE}
nypd_year_vic_male <- nypd %>%
    filter(VIC_SEX=='M') %>%
    mutate(year = format(datetime, "%Y")) %>%
    group_by(year) %>%
    summarise(vic_male_count=n()) %>%
    select(year, vic_male_count) %>%
    ungroup()
nypd_year_vic_male
```
```{r join_year_data, echo=TRUE}
nypd_year_stat <- nypd_year %>%
    full_join(nypd_year_vic_male)
nypd_year_stat
```

2) Create the linear model

```{r create_model, echo=TRUE}
mod <- lm(vic_male_count ~ count, data=nypd_year_stat)
summary(mod)
```

3) Predict and plot for male victim incidents
```{r predit_male, echo=TRUE}
male_pred <- nypd_year_stat %>% mutate(pred = predict(mod))

male_pred %>% ggplot() +
    geom_point(aes(x=count, y=vic_male_count), color="blue") +
    geom_point(aes(x=count, y=pred), color="red") +
    geom_line(aes(x=count, y=pred), color="red")
```

As you can observe, the actual data aligns very well with this model. Not only in
average, male incidents represented 90% of the total cases, but also it approached
90% consistently across each year.

### Conclusion:

From 2006 to 2019, the situation regarding shooting incidents showed improvement
year by year. However, with the onset of the COVID-19 pandemic in 2020, the number
nearly doubled compared to the previous year, and it has remained at that elevated
level in subsequent years. Upon conducting some research, it appears that increased
firearm sales and deteriorating mental health in individuals are contributing to
the spike in cases. It is imperative for New York City to address these issues
promptly.

Moreover, a significant proportion of these incidents occurred during the night
and outdoors. Therefore, individuals are advised to refrain from going out in the
late evening to ensure their safety.

Among all the boroughs, Brooklyn recorded the highest number of shooting incidents.

### Possible bias:

When conducting statistical analysis on the incidents in each borough, population
was not taken into account. While Brooklyn had the highest number of shooting
incidents, it also had the largest population. If population was factored in, it
appears that the Bronx had the highest rate of shooting incidents per capita.


```{r sessioninfo, echo=TRUE}
sessionInfo()
```
