---
title: "NYPD Shooting incident analysis"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import the NYPD shooting incident data


```{r import_csv, echo=TRUE}
library(tidyverse)
library(lubridate)
filename = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd <- read_csv(filename)
nypd
```

Merge OCCUR_DATE and OCCUR_TIME into "dttm" format
```{r merge_date_time, echo=TRUE}
nypd <- nypd %>%
    unite(datetime, 2:3, sep = " ", remove = FALSE) %>%
    mutate(
        datetime = as.POSIXct(datetime, format = "%m/%d/%Y %H:%M:%S")) %>%
    ungroup()
nypd %>%
    select(datetime, OCCUR_DATE, OCCUR_TIME, everything())
```

#### Get insight per Borough


```{r get_boro_stat, echo=TRUE}
boro_stat <- nypd %>%
    group_by(BORO) %>%
    summarise(count=n()) %>%
    select(BORO, count) %>%
    ungroup()
boro_stat
```

```{r draw_boro_stat_pie, echo=TRUE}
slices <- boro_stat$count
boro <- boro_stat$BORO
percent <- round(slices/sum(slices)*100)
boro <- paste(boro, percent)
boro <- paste(boro, "%", sep="")
pie(slices, labels=boro, main="Pie Chart of Boroughs")
```

As we can wee, Brooklyn has the the most shooting incidents (40%), followed by
Bornx (29%).


#### Check the incidents over the years


```{r generate_year_stat, echo=TRUE}
nypd <- nypd %>% mutate(OCCUR_DATE=mdy(OCCUR_DATE))

nypd_year <- nypd %>%
    group_by(datefloor=lubridate::floor_date(OCCUR_DATE, "year")) %>%
    summarise(count=n()) %>%
    mutate(year=format(datefloor, format="%Y")) %>%
    select(year, count) %>%
    ungroup()
nypd_year
```
```{r draw_year_stat, echo=TRUE}
nypd_year %>%
  ggplot(aes(x=year, y=count, fill=count)) +
  geom_bar(width=0.8, stat="identity") +
  labs(title = "NY shooting incidents over the years", y = NULL)
```

During 2017-2019, the shooting incidents per year were below 1000. But incidents
jumped to nearly doubled (close to 2000) in year 2020.

#### Check incidents in each hour


```{r generate_hour_stat, echo=TRUE}
nypd_hour_stat <- nypd %>%
    mutate(hour = format(datetime, "%H")) %>%
    group_by(hour) %>%
    summarise(count=n()) %>%
    select(hour, count) %>%
    ungroup()
print(nypd_hour_stat, n = Inf)
```

```{r draw_hour_stat, echo=TRUE}
nypd_hour_stat %>%
  ggplot(aes(x=hour, y=count, fill=count)) +
  geom_bar(width=0.8, stat="identity") +
  labs(title = "NY shooting incidents over each hour of the day", y = NULL)
```

As showed above, the peak time of shooting incidents was between 22:00 ~ 02:00.
The hour with the highest incidents was 23:00 - 00:00.

#### Check the class of location


```{r check_loc_class_data, echo=TRUE}
nypd %>%
    group_by(LOC_CLASSFCTN_DESC) %>%
    summarise(count=n()) %>%
    select(LOC_CLASSFCTN_DESC, count) %>%
    ungroup()
```

Seems we have lots of NAs... drop them first

```{r get_loc_class_data, echo=TRUE}
nypd_loc_class <- nypd %>%
    drop_na(LOC_CLASSFCTN_DESC) %>%
    group_by(LOC_CLASSFCTN_DESC) %>%
    summarise(count=n()) %>%
    select(LOC_CLASSFCTN_DESC, count) %>%
    ungroup()
nypd_loc_class
```

```{r draw_loc_class_pie, echo=TRUE}
slices <- nypd_loc_class$count
loc_class <- nypd_loc_class$LOC_CLASSFCTN_DESC
percent <- round(slices/sum(slices)*100)
loc_class <- paste(loc_class, percent)
loc_class <- paste(loc_class, "%", sep="")
pie(slices, labels=loc_class, main="Pie Chart of Location class", cex=0.8)
```

More than half the incidents happened in the street!

#### Check Inside and Outside incidents


```{r check_in_out_state, echo=TRUE}
nypd %>%
     drop_na(LOC_OF_OCCUR_DESC) %>%
     group_by(LOC_OF_OCCUR_DESC) %>%
     summarise(count=n()) %>%
     select(LOC_OF_OCCUR_DESC, count) %>%
     ungroup()
```

Outside is the majority case.

#### Check the victim gender data


```{r get_victim_gender, echo=TRUE}
nypd_victim_gender <- nypd %>%
    group_by(VIC_SEX) %>%
    summarise(count=n()) %>%
    select(VIC_SEX, count) %>%
    ungroup()
nypd_victim_gender
```

```{r draw_victim_gendor_pie, echo=TRUE}
slices <- nypd_victim_gender$count
vic_sex <- nypd_victim_gender$VIC_SEX
percent <- round(slices/sum(slices)*100)
vic_sex <- paste(vic_sex, percent)
vic_sex <- paste(vic_sex, "%", sep="")
pie(slices, labels=vic_sex, main="Pie Chart of the gender of the victim")
```

As showed above, Male as victim dominates the cases(90%).

```{r sessioninfo, echo=TRUE}
sessionInfo()
```
